Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-06-16 manuscript/manuscript.Rmd
adds to discussion

Discussion
----------
We sought to test the assumption that the sensitivities of plant populations to climate variables can be detected equally well using either individual level data or population level data. This is an important question to answer because population models are key tools for predicting the consequences of global climate change. However, they can be of limited use when built on data from a small subset of a population in space or time. If population level data (i.e., some aggregated form of individual level data) can be used to detect climate effects on population dynamics, then we would have a cheaper and easier option for data collection over relatively large temporal and spatial extents [_e.g._ @Freckleton2011].

Our analysis offers good news and bad news. The good news is that for two out of our four focal species, the QBM faired no worse than the demographically detailed IPM in terms of mean absolute (Table 1). Thus, it seems that aggregate level data can provide forecasts of similar skill as individual level data.

>>>>

Discussion
----------
Demographic models are costly to parameterize, yet remain the most widely-used tools for studying plant population dynamics (e.g., @Salguero-Gomez2015). New density-structured approaches sacrifice some precision at the local scale for better spatial coverage that is virtually unattainable, due to time and funding constraints, using traditional individual-based data [@Freckleton2011; @Queenborough2011]. These approaches are appealing because traditional population models suffer from constrained parameter space -- meaning forecasts quickly extend beyond the parameter space used to fit the model. Density-structured approaches rely on easy-to-collect census data (e.g., percent cover in $1\text{m}^2$ quadrats) so sampling populations across large spatial extents would be possible, thus covering more potential parameter space.

Estimating parameter variability in space and time is critical when using population models to forecast responses to directional changes in exogoneous drivers like climate [@Clark2011; @Vanderwel2013; @Wolkovich2014a; @Petchey2015]. Compared to demographic models, density-structured modeling approaches may provide an efficient route toward estimating such uncertainty. However, population models based on aggregated inidividual data have yet to be tested with climate drivers. Since it is individuals that respond to weather, rather than populations, models based on a aggregated data may produce incorrect forecasts [@Clark2012]. Thus, we sought to compare climate sensitivities of individual and population based models of four grassland species populations.

### 2015-06-16 manuscript/manuscript.tex
adds to discussion

We use an environmentally stochastic IPM (Rees and Ellner 2009) that
includes the random year effects and the climate covariates from the
vital rate statistical models. But note that we can, and do for some
simulations, ignore the random year effects so that only the climate
effects can drive interannual variation. Our IPM follows the
specification of Chu and Adler (2015) where the population of species
\emph{j} is a density function $n(u_{j},t)$ giving the density of
sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
scale, so that $n(u_{j},t)du$ is the number of genets whose area (on the
arithmetic scale) is between $e^{u_{j}}$ and $e^{u_{j}+du}$. So, the
density function for any size \emph{v} at time $t+1$ is

@@

We sought to test the assumption that the sensitivities of plant
populations to climate variables can be detected equally well using
either individual level data or population level data. This is an
important question to answer because population models are key tools for
predicting the consequences of global climate change. However, they can
be of limited use when built on data from a small subset of a population
in space or time. If population level data (i.e., some aggregated form
of individual level data) can be used to detect climate effects on
population dynamics, then we would have a cheaper and easier option for
data collection over relatively large temporal and spatial extents
(\emph{e.g.} Freckleton et al. 2011).

Our analysis offers good news and bad news. The good news is that for
two out of our four focal species, the QBM faired no worse than the
demographically detailed IPM in terms of mean absolute (Table 1). Thus,
it seems that aggregate level data can provide forecasts of similar
skill as individual level data.

@@

\begin{table}[ht]
\centering
\caption{Accuracy (mean absoluet error, MAE) and precision (90\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

\emph{NOTES}: The IPM MAE is significantly lower at $\alpha=0.05$ for
\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
are statisticially similar between models for \emph{P. secunda}
(\emph{P} = 0.0922).

@@

@@

@@

Rees, M., and S. P. Ellner. 2009. Integral projection models for
populations in temporally varying environments. Ecological Monographs
79:575--594.

@@

>>>>

We use an environmentally stochastic IPM (Mark and Ellner 2009) that
includes the random year effects and the climate covariates from the
vital rate statistical models. But note that we can, and do for some
simulations, ignore the random year effects so that only the climate
effects can drive interannual variation. Our IPM follows the
specification of Chu and Adler (2015) where the population of species
\emph{j} is a density function \(n(u_{j},t)\) giving the density of
sized-\emph{u} genets at time \emph{t}. Genet size is on the natural log
scale, so that \(n(u_{j},t)du\) is the number of genets whose area (on
the arithmetic scale) is between \(e^{u_{j}}\) and \(e^{u_{j}+du}\). So,
the density function for any size \emph{v} at time \(t+1\) is

@@

Demographic models are costly to parameterize, yet remain the most
widely-used tools for studying plant population dynamics (e.g.,
Salguero-Gómez et al. (2015)). New density-structured approaches
sacrifice some precision at the local scale for better spatial coverage
that is virtually unattainable, due to time and funding constraints,
using traditional individual-based data (Freckleton et al. 2011,
Queenborough et al. 2011). These approaches are appealing because
traditional population models suffer from constrained parameter space --
meaning forecasts quickly extend beyond the parameter space used to fit
the model. Density-structured approaches rely on easy-to-collect census
data (e.g., percent cover in \(1\text{m}^2\) quadrats) so sampling
populations across large spatial extents would be possible, thus
covering more potential parameter space.

Estimating parameter variability in space and time is critical when
using population models to forecast responses to directional changes in
exogoneous drivers like climate (Clark et al. 2011, Vanderwel et al.
2013, Wolkovich et al. 2014, Petchey et al. 2015). Compared to
demographic models, density-structured modeling approaches may provide
an efficient route toward estimating such uncertainty. However,
population models based on aggregated inidividual data have yet to be
tested with climate drivers. Since it is individuals that respond to
weather, rather than populations, models based on a aggregated data may
produce incorrect forecasts (Clark et al. 2012). Thus, we sought to
compare climate sensitivities of individual and population based models
of four grassland species populations.

@@

\begin{table}[ht]
\centering
\caption{Accuracy (mean absolute error, MAE) and precision (90\% Distance) of out of sample predictions. Forecasts were made without random year effects; only climate covariates could explain year-to-year variation. 90\% Distance refers to the average distance between the upper and lower 90th percentiles of the 100 predicted values for each quadrat-year combination.} 
\begin{tabular}{llrrr}
  \hline
Species & Model & MAE & 90\% Distance & Mean Obs. Cover \\ 
  \hline
BOGR & IPM & 12.18 & 38.52 & 9.43 \\ 
  BOGR & QBM & 19.66 & 56.50 & 9.26 \\ 
  HECO & IPM & 1.22 & 6.47 & 1.15 \\ 
  HECO & QBM & 12.35 & 41.11 & 1.18 \\ 
  PASM & IPM & 0.19 & 1.65 & 0.42 \\ 
  PASM & QBM & 0.55 & 7.78 & 0.42 \\ 
  POSE & IPM & 1.37 & 7.64 & 1.25 \\ 
  POSE & QBM & 1.79 & 40.59 & 1.27 \\ 
   \hline
\end{tabular}
\end{table}

\emph{NOTES}: The IPM MAE is significantly lower at \(\alpha=0.05\) for
\emph{B. gracilis} (\emph{P} = 0.0012), \emph{H. comata} (\emph{P} =
4.0586 × 10-8), and \emph{P. smithii} (\emph{P} = 3.183 × 10-5). MAEs
are statisticially similar between models for \emph{P. secunda}
(\emph{P} = 0.0922).

@@

Clark, J. S., D. M. Bell, M. H. Hersh, and L. Nichols. 2011. Climate
change vulnerability of forest biodiversity: Climate and competition
tracking of demographic rates. Global Change Biology 17:1834--1849.

@@

Mark, R., and S. P. Ellner. 2009. Integral projection models for
populations in temporally varying environments. Ecological Monographs
79:575--594.

Petchey, O. L., M. Pontarp, T. M. Massie, S. Kéfi, A. Ozgul, M.
Weilenmann, G. M. Palamara, F. Altermatt, B. Matthews, J. M. Levine, D.
Z. Childs, B. J. McGill, M. E. Schaepman, B. Schmid, P. Spaak, A. P.
Beckerman, F. Pennekamp, and I. S. Pearse. 2015. The ecological forecast
horizon, and examples of its uses and determinants. Ecology Letters
18:597--611.

@@

@@

Vanderwel, M. C., V. S. Lyutsarev, and D. W. Purves. 2013.
Climate-related variation in mortality and recruitment determine
regional forest-type distributions. Global Ecology and Biogeography
22:1192--1203.

Wolkovich, E. M., B. I. Cook, K. K. McLauchlan, and T. J. Davies. 2014.
Temporal ecology in the Anthropocene. Ecology Letters 17:1365--1379.
